---
title: "Covid-19 data analysis"
author: "Raj S."
date: "2023-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Covid-19 Data Analysis

```{r}
#Loading the tidyverse library
library(tidyverse)
```

```{r}
# Get the current data in all the four files 
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
```

Reading into the and reviewing it.

```{r import_data, message = FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])

global_cases
```

Now on reviewing the global cases data, we can see that we do not the Lay and Log data, while at the same time we need the dates to in each row rather than each column.

```{r tidy_global_data}
global_cases <- global_cases %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "cases" ) %>%
  select(-c(Lat, Long))

global_cases
```

Now lets review the global_deaths data 

```{r global_deaths}
global_deaths
```

global_deaths has the same format of data in terms of columns as global_cases and thus we will perform similar operation on it, as we did on the global_deaths data

```{r tidy_global_data}
global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), names_to = "date", values_to = "deaths" ) %>%
  select(-c(Lat, Long))

global_deaths
```

We will be now joining both the datasets - global_cases & global_deaths into a single data frame and also covert the date fro character object to date object.

```{r, global}
library(lubridate)
global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>% 
    mutate(date = mdy(date))
  
global
summary(global)
```

Now there are rows where the number of cases are zero and thus it has no value to us for type of analysis at this point.

```{r}
global %>% filter(cases>0)
summary(global)
```
Now lets check the validity of data specifically in terms of number of cases and confirm that the given number is not a typo.

```{r}
global %>% filter(cases>100000000)
```

Thus, we can confirm that the number of cases data is valid and correct.